Schema Change: Challenge Problems

### 2024-05-21 paper.tex
Add TP rough sketch paragraph

>>>>

\section{\textcolor{red}{[TP] Rough paragraph}}

Programming is increasingly collaborative and interactive. We welcome this trend, because
greater collaboration and interactivity make programming more effective. Future programming
environments enable programmers to collaborate at a more fine-grained level, adopting individual
code changes created by their collaborators as needed. They also let programmers interactively
edit code, often during program execution.

In order to support this new way of working, programming systems increasingly have to tackle the
problem of schema evolution. If a programmer imports a change that edits a schema, code and data
that relies on such schema needs to be adapted to match the new schema. Similarly, if a programmer
in a live programming environment modifies their type definition, current program state needs to
be updated to match the new structure of types. This is a well-understood and well-studied problem
in database community, but it has received little attention in work on programming systems.

The aim of this paper is to provide a conceptual framework for talking about schema evolution in
collaborative and live programming systems. Taking inspiration from Stewart Brand's analysis of
pace layers, we propose to think of the problem as the problem of evolution of occuring at three
layers, the most permanent schema layer, less permanent code layer and least permanent data layer.
A change at the most permanent layer affects the less permanent layer, which then has to be
updated to keep correspondence between the layers.

We consider four case studies of such schema evolution in collaborative and live programming
systems. We present those throught the perspective of our unified conceptual framework. The work
both sheds a new light at how schema change needs to be handled in programming systems and also
provides a range of challenges that show interesting problems that future programming systems
will need to tackle.
